---
title: "alpha diversity of marine viral clusters"
author: "Xiaotian"
date: "2021/4/8"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setting environment

```{r}
rm(list = ls())
library("vegan")
library("ggplot2")
library("reshape2")
library("tidyverse")
library("patchwork")
setwd("C:/Users/sherry/Documents/GitHub/QBVirus")
#load data
vcs.o <-read.csv("./data/vcs.csv", row.names = 1)
dim(vcs.o)
vcs <- as.data.frame(t(vcs.o))
```

## Alpha diversity metrics

Considering the decimal abundance based on contig-coverage, we apply number of species, Shannon index and Smith & Wilson's Evenness to estimate richness (S), alpha-diversity (H) and evenness (E)of each site, separately.

```{r}
# richness
S.obs <- function( x = "" ){
  rowSums( x > 0 ) * 1
}
# evenness
Evar <- function(x){
  x <- as.vector(x[x > 0])
  1 - (2/pi)*atan(var(log(x)))
}
# ShannonH
ShanH <- function(x = ""){
  H = 0
  for (n_i in x){
    if(n_i > 0) {
      p = n_i / sum(x)
      H = H - p*log(p)
    }
  }
  return(H)
}
# calculating
vcs.S <- c()
vcs.E <- c()
vcs.H <- c()
for (i in 1:91) {
  vcs.S[i] = S.obs(vcs[i,])
  vcs.E[i] = Evar(vcs[i,])
  vcs.H[i] = ShanH(vcs[i,])
}
vcs.div <- cbind(vcs.S, vcs.E, vcs.H)
write.csv("./data/vcs_div.csv")
```

## Stacked barplots with different grouping strategies 

```{r}
#convert to relative abundance table
vcs.rel <- vcs/rowSums(vcs)
# top20 VCs
top <- names(head(sort(colSums(vcs),decreasing = T),20))
top <- as.vector(top)
# recreate a table with the rest VCs classified into "Others"
vcs.rel2 <- vcs.rel
colnames(vcs.rel2)[!colnames(vcs.rel2)%in%top] <- "Others"
others <- rowSums(cbind(vcs.rel2[colnames(vcs.rel2)=="Others"]))
vcs.top <- cbind(vcs.rel2[colnames(vcs.rel2)!="Others"],others)
vcs.top <- vcs.top[,order(colSums(vcs.top))] # ranking
#load group data
env.class.geo.div <- read.table("C:/Users/sherry/Documents/GitHub/QBVirus/data/env-geo-div-class.txt",
                                sep = "\t", header = TRUE,row.names = 1)
vcs.top2 <- as.data.frame(cbind(vcs.top,env.class.geo.div$Layer,env.class.geo.div$Region))
vcs.top2$sample <- rownames(vcs.top2)
vcs.top20 <- melt(vcs.top2,ID="names")
colnames(vcs.top20)[names(vcs.top20)=="variable"]<-"Taxa"
vcs.top20$`env.class.geo.div$Layer` <- factor(vcs.top20$`env.class.geo.div$Layer`, levels = c("SUR","DCM","MIX","MES")) #set order
# group by layer(zonation)
vcs.top.layer <- aggregate(vcs.top2[,1:21], by = list(Layer = vcs.top2$`env.class.geo.div$Layer`),FUN = mean)
vcs.top.layer$sample <- rownames(vcs.top.layer)
vcs.group.layer <- melt(vcs.top.layer, ID="names")
colnames(vcs.group.layer)[names(vcs.group.layer)=="variable"]<-"Taxa"
vcs.group.layer$Layer <- factor(vcs.group.layer$Layer, levels = c("SUR","DCM","MIX","MES")) #set order
# group by region
vcs.top.region <- aggregate(vcs.top2[,1:21], by = list(Layer = vcs.top2$`env.class.geo.div$Region`),FUN = mean)
vcs.top.region$sample <- rownames(vcs.top.region)
vcs.group.region <- melt(vcs.top.region, ID="names")
colnames(vcs.group.region)[names(vcs.group.region)=="variable"]<-"Taxa"
```

```{r}
#stacked barplot for top 20 viral clusters
colors<-c("grey50","darkolivegreen3","gold","darkseagreen","chartreuse4","darkorange","dodgerblue4","burlywood2","brown3","#984EA3","cyan3","#CC6600","darkgrey","3300CC","lightgrey","pink","yellow","white","000099","66CC66","darkgreen")
ggplot(vcs.top20,aes(x = sample, y = value, fill = Taxa))+
  geom_bar(position = "fill", stat = "identity", width = 1)+
  theme_bw()+
  scale_fill_manual(values=colors)+
  facet_wrap(~env.class.geo.div$Layer,nrow = 4)+
  labs(x = "Sites",y = "Relative Abundance", fill = "Viral clusters")+
  theme(axis.text.x = element_text(size = 0, color = "transparent"),
        legend.text = element_text(size = 7))+
  guides(fill = guide_legend(keywidth = 0.5, keyheight = 0.5))
```

```{r}
# group by water layer
ggplot(vcs.group.layer,aes(x = Layer, y = value, fill = Taxa))+
  geom_bar(position = "fill", stat = "identity")+
  theme_bw()+
  scale_fill_manual(values=colors)+
  labs(x = "Zone",y = "Relative Abundance", fill = "Viral clusters")+
  theme(axis.text.x = element_text(size = 9, color = "black"),
        legend.text = element_text(size = 7))+
  guides(fill = guide_legend(keywidth = 0.5, keyheight = 0.5))
# group by region
ggplot(vcs.group.region,aes(x = Layer, y = value, fill = Taxa))+
  geom_bar(position = "fill", stat = "identity")+
  theme_bw()+
  scale_fill_manual(values=colors)+
  labs(x = "Region",y = "Relative Abundance", fill = "Viral clusters")+
  theme(axis.text.x = element_text(size = 9, color = "black"),
        legend.text = element_text(size = 7))+
  guides(fill = guide_legend(keywidth = 0.5, keyheight = 0.5))
```

## Seeking relationships with diversity and water depth
```{r}
# linear regression
summary(lm(env.class.geo.div$ShannonH ~ env.class.geo.div$Depth))
summary(lm(env.class.geo.div$Richness ~ env.class.geo.div$Depth))
summary(lm(env.class.geo.div$Evenness ~ env.class.geo.div$Depth))

HD <- ggplot(env.class.geo.div, aes(x = Depth, y = ShannonH))+
  geom_point()+
  geom_smooth(method = "lm", alpha = 0.2)
SD <- ggplot(env.class.geo.div, aes(x = Depth, y = Richness))+
  geom_point()+
  geom_smooth(method = "lm", alpha = 0.2)
ED <- ggplot(env.class.geo.div, aes(x = Depth, y = Evenness))+
  geom_point()+
  geom_smooth(method = "lm", alpha = 0.2)
HD + SD + ED
```